\Title

### 2020-12-15 manuscript/manuscript.tex
Apply Santiago's comments about defining some parameters
Co-authored-by: Santiago Soler <santiago.r.soler@gmail.com>

\subsection{Damped least-squares solution}
\label{sec:eql-inversion}

We can obtain the values of the source coefficients $\mathbf{c}$ that best
fit the observed field values $\mathbf{d}^o$ by minimizing the goal function

\begin{equation}
    \phi(\mathbf{c}) =
    \left[\mathbf{d}^o - \mathbf{A}\mathbf{c}\right]\trans
    \mathbf{W}
    \left[\mathbf{d}^o - \mathbf{A}\mathbf{c}\right]
    + \lambda_d\ \mathbf{c}\trans\mathbf{c}
    \ ,
    \label{eq:misfit-unscaled}
\end{equation}

\noindent where
$\mathbf{W}$ is a $N \times N$ diagonal matrix of data weights and
$\lambda_d$ is a positive \emph{damping} parameter with the same units as the
Jacobian matrix elements.
The second term on the right-hand side of the equation is a zeroth-order
Tikhonov regularization \citep{tikhonov1977} (also known as a damping
regularization or ridge regression) that is used to stabilize the solution.

The range of acceptable values for the damping parameter $\lambda_d$ will
depend on the values of the Jacobian matrix $\mathbf{A}$ and the coefficients.
Consequently, this range will vary (often dramatically) between datasets,
making it difficult to choose an appropriate value for $\lambda_d$ in practice.

To solve this issue, we first scale the Jacobian matrix so that its elements
are unitless and each column has unit variance.
We define a diagonal matrix $\mathbf{S}$

\begin{equation}
    \mathbf{S} =
    \begin{bmatrix}
      \sigma_1 & 0 & \cdots &0 \\
      0 & \sigma_2 & \cdots &0 \\
      \vdots & \vdots & \ddots & \vdots \\
      0  & 0 & \cdots & \sigma_M
    \end{bmatrix}_{M \times M}
    ,
\end{equation}

\noindent in which $\sigma_j$ is the standard deviation of the $j$-th column of
$\mathbf{A}$.
We then write the forward problem in equation~\ref{eq:linear-problem} as

\begin{equation}
    \mathbf{d}
    =
    \mathbf{A} \mathbf{S}\inv \mathbf{S} \mathbf{c}
    =
    \left[
        \mathbf{A} \mathbf{S}\inv
    \right]
    \left[
        \mathbf{S} \mathbf{c}
    \right]
    =
    \mathbf{B} \mathbf{m}
\end{equation}

\noindent where $\mathbf{B}$ is the scaled and unitless Jacobian matrix and
$\mathbf{m}$ is a vector containing scaled coefficients with the same units as
the data.

The goal function defined in equation~\ref{eq:misfit-unscaled} can be
rewritten as

\begin{equation}
    \phi(\mathbf{m}) =
    \left[\mathbf{d}^o - \mathbf{B}\mathbf{m}\right]\trans
    \mathbf{W}
    \left[\mathbf{d}^o - \mathbf{B}\mathbf{m}\right]
    + \lambda\ \mathbf{m}\trans\mathbf{m}
    \ ,
    \label{eq:misfit}
\end{equation}

\noindent where $\lambda$ is a \emph{dimensionless} damping parameter and
regularization is applied on the scaled coefficients $\mathbf{m}$ instead of
$\mathbf{c}$.

The vector of scaled coefficients $\hat{\mathbf{m}}$ that minimizes the goal
function can be found by solving the \emph{normal equation system}
\citep{menke1989}

\begin{equation}
    \left[
      \mathbf{B}\trans \mathbf{W} \mathbf{B} + \lambda \mathbf{I}
    \right]
    \hat{\mathbf{m}} =
    \mathbf{B}\trans\mathbf{W}
    \mathbf{d}^o.
\end{equation}

Once the scaled coefficients are obtained, the estimated unscaled coefficients
$\hat{\mathbf{c}}$ can be calculated by removing the scaling factor

\begin{equation}
    \hat{\mathbf{c}} = \mathbf{S}\inv \hat{\mathbf{m}} \ .
\end{equation}

\noindent The forward modeling operations used to perform predictions
(e.g., for interpolation and upward continuation) are left unchanged by
using vector $\hat{\mathbf{c}}$ instead of $\hat{\mathbf{m}}$.

@@

\begin{table*}
    \centering
    \caption{
        Source distributions as combinations of source layouts and depth
        strategies.
    }
    \label{tab:source-distributions}
    \begin{tabular}{lccc}
        & Source below data & Block averaged sources & Grid sources \\ \hline
        Constant depth & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
        Relative depth & $\checkmark$ & $\checkmark$ & $\times$     \\
        Variable depth & $\checkmark$ & $\checkmark$ & $\times$     \\
    \end{tabular}
\end{table*}

\subsection{Gradient boosted equivalent sources}

Reducing the number of sources might help to lower the amount of computer
memory needed to fit their coefficients, although it may be not enough for
a very large set of observation points.
Even if we can greatly reduce the number of sources, having too many data
points would still produce a Jacobian matrix large enough that cannot fit in
memory.
This problem was already faced by other authors
\citep{mendonca1994, leao1989, li2010, barnes2011, oliveira2013, siqueira2017,
jirigalatu2019}.
The iterative method proposed by \citet{siqueira2017} needs equivalent sources
to be located beneath each observation point, so it cannot take advantage of
the block-averaged sources.
\citet{leao1989} perform the fitting simultaneously with the prediction, what
prevents us to use already fitted equivalent sources to make future
predictions. Besides the method is limited to interpolations on regular grids.
% Add more discussion on why we won't apply any of the following methods
\citet{mendonca1994}
\citet{li2010}
\citet{barnes2011}
\citet{jirigalatu2019}

Here we propose an iterative algorithm for fitting the source coefficients
using a gradient boosting \citep{friedman2001, friedman2002} strategy:
(i) define a collection of overlapping windows of equal size that cover the
survey area,
(ii) define one set of equivalent sources per window,
(iii) iteratively fit the coefficients of each set of sources.
Once every set of sources is fitted, we can predict the value of the field on
any observation point as the effect of every set of equivalent sources.

Lets assume we have $N$ measurements of a harmonic field on observations points
$\{\mathbf{p}_i\}_{i=1}^{N}$ stored in the vector $\mathbf{d} = (d_1, d_2,
\ldots, d_N)\trans$. We have defined a set of $M$ point sources located on
$\{\mathbf{q}_j\}_{j=1}^{M}$. Instead of trying to fit the entire set of
coefficients $\mathbf{c}$ at once, we can reduce the problem to fitting $K$
different sets of equivalent sources: we define $K$ overlapping windows of
equal size that cover the whole observation area, and then we create one set of
equivalent sources per window formed by the $\{\mathbf{q}_m\}_{m=1}^{M_k}$
sources that fall inside the $k$-th window.
The observed data should be equal to the superposition of the
effect of every set of equivalent sources:

\begin{equation}
    \mathbf{d} = \sum_{k=1}^K \alpha_k \mathbf{A}_k \mathbf{c}_k,
\end{equation}

\noindent where $\mathbf{c}_k$ is the vector of coefficients for the $k$-th
set of equivalent sources; $\mathbf{A}_k$ is the Jacobian matrix between
those sources and every data point; and each $\alpha_k$ is a constant called
\emph{step-size parameter}.

By applying gradient boosting we can fit each $\mathbf{c}_k$ vector
iteratively: on each iteration a residue vector $\mathbf{r}_k$ (initialized
with the observed data) is used to fit the $\mathbf{c}_k$
coefficients and $\alpha_k$ step-size parameter, and then updated by removing
the effect of the same sources, as follows:

\begin{equation}
    \mathbf{r}_0 = \mathbf{d},
\end{equation}

\begin{equation}
    \mathbf{r}_k = \mathbf{r}_{k-1} - \alpha_k \mathbf{A}_k\mathbf{c}_k.
    \label{eq:residue}
\end{equation}

\noindent On each iteration the $\mathbf{c}_k$ vector and the $\alpha_k$
parameter can be fitted so they minimize the misfit function:

\begin{equation}
    \phi(\mathbf{c}_k) =
        \left\lVert
            \mathbf{r}_{k-1} - \alpha_k \mathbf{A}_k\mathbf{c}_k
        \right\rVert ^ 2
        + \lambda_d \left\lVert \mathbf{c}_k \right\rVert ^2,
\end{equation}

\noindent following the same method described on Section
\ref{sec:eql-inversion}.
Because the $\mathbf{A}_k$ matrices have only $N \times M_k$ elements,
adjusting the size of the overlapping windows can help in reducing the computer
memory needed to fit the source coefficients.
Nevertheless, in case we have too many data points that make the $\mathbf{A}_k$
matrices large enough that they cannot be stored in memory, we can apply one
more simplification on the fitting process: ignore the effect of the $k$-th
sources on any observation point that falls outside the $k$-th window.

Assuming that there are $N_k$ observation points that fall under the $k$-th
window, we can fit the $\mathbf{c}_k$ coefficients by minimizing the following
misfit function:

\begin{equation}
    \phi(\mathbf{c}_k) =
        \left\lVert
            \tilde{\mathbf{r}}_{k-1} - \tilde{\mathbf{A}}_k\mathbf{c}_k
        \right\rVert ^ 2
        + \lambda_d \left\lVert \mathbf{c}_k \right\rVert ^2,
\end{equation}

\noindent where $\tilde{\mathbf{r}}_{k-1}$ is the residue vector containing
only the elements that correspond to the $\{\mathbf{p}_n\}_{n=1}^{N_k}$
observation points that fall under the $k$-th window; and
$\tilde{\mathbf{A}}_k$ is the Jacobian matrix between these observation points
and the $k$-th sources.
Because the $\tilde{\mathbf{A}}_k$ matrices will have $N_k \times M_k$
elements, the computer memory needed for fitting the $\mathbf{c}_k$
coefficients is greatly reduced.

Once the $\mathbf{c}_k$ coefficients are fitted following the same method
described on Section \ref{sec:eql-inversion}, we need to obtain the value
of the step-size parameter $\alpha_k$.
This can be achieved by minimizing the norm of the difference between the
previous residue $\mathbf{r}_{k-1}$ and the effect of the $k$-th sources on
every observation point:

\begin{equation}
    \alpha_k = \underset{\alpha}{\mathrm{argmin}}
    \left(
        \left\lVert
            \mathbf{r}_{k-1} - \alpha \mathbf{A}_k \mathbf{c}_k
        \right\rVert ^ 2
    \right),
\end{equation}

\noindent which leads to:

\begin{equation}
    \alpha_k = \frac{
        \mathbf{r}_{k-1}\trans \mathbf{A}_k \mathbf{c}_k
    }{
        \left(\mathbf{A}_k \mathbf{c}_k\right)\trans \mathbf{A}_k \mathbf{c}_k
    }.
\end{equation}

This step will improve the convergence of the iterative algorithm and also will
introduce at some extent the effect of the $k$-th sources on every data point,
which was neglected on the fitting of the $\mathbf{c}_k$ coefficients.

Once both the $\alpha_k$ parameter and the $\mathbf{c}_k$ coefficients are
fitted, equation \ref{eq:residue} is applied to update the residue: compute the
effect of the sources belonging to the $k$-th window on every observation
point. Worth noting that when calculating the effect of any number of sources
we don't need to pre-compute the entire Jacobian matrix: we can calculate the
effect of the sources on each observation point by a simple addition of terms,
what means computing each Jacobian element at once when needed.

Because the number of windows is fixed before starting the iteration process,
its number of steps are also fixed to $K$. Moreover, the windows are randomly
shuffled before starting the process to prevent the prioritization of clustered
sets of equivalent sources.

After all $\alpha_k$ parameters and $\mathbf{c}_k$ coefficients vectors are
fitted, we can predict the effect of every equivalent source on any point as
follows:

\begin{equation}
    V(\mathbf{p}) =
    \sum\limits_{k=1}^K \sum\limits_{m=1}^{M_k}
    \frac{\alpha_k {c_k}_m}{|\mathbf{p} - \mathbf{q}_m|}.
\end{equation}

>>>>

\subsection{Fitting source coefficients}
\label{sec:eql-inversion}

Solving equations~\ref{eq:linear-problem-g} and~\ref{eq:linear-problem} is
equivalent to performing an inversion and thus solving an ill-posed problem by
nature due to the non-uniqueness and the instability of the possible solutions.
Therefore is usually needed to add regularization functions in order to obtain
stable solutions
(\citealp[p.~52]{menke1989}; \citealp{cooper2000}; \citealp{oliveira2013}).

We can obtain the values of the source coefficients $\mathbf{c}$ that better
reproduce the measured field by minimizing an objective or misfit function
$\phi(\mathbf{c})$ through a least squares method:

\begin{equation}
    \phi(\mathbf{c}) =
        \left\lVert
            \mathbf{d} - \mathbf{A}\mathbf{c}
        \right\rVert ^ 2
        + \lambda_d \left\lVert \mathbf{c} \right\rVert ^2,
    \label{eq:misfit-unscaled}
\end{equation}

\noindent where $\left\lVert \cdot \right\rVert ^ 2$ symbolizes the L$_2$ norm
and $\lambda_d$ is a \emph{damping} parameter with the same units as the
Jacobian matrix elements.
The second term of the equation applies a zeroth-order Tikhonov
\citep{tikhonov1977} regularization to stabilize the solution.

In order to simplify the choice of the \emph{damping} parameter we scale the
Jacobian matrix so each column has unit variance.
We define a diagonal matrix $\mathbf{S}$, whose elements corresponds to the
standard deviation of each column of the Jacobian matrix $\mathbf{A}$:

\begin{equation}
    \mathbf{S} =
    \begin{bmatrix}
        \sigma_1 & \cdots &0 \\
        \vdots & \ddots & \vdots \\
        0 & \cdots & \sigma_M
    \end{bmatrix}
    ,
\end{equation}

\noindent so equation~\ref{eq:linear-problem} can be written as:

\begin{equation}
    \mathbf{d}
    =
    \mathbf{A} \mathbf{S}\inv \mathbf{S} \mathbf{c}
    =
    \left[
        \mathbf{A} \mathbf{S}\inv
    \right]
    \left[
        \mathbf{S} \mathbf{c}
    \right]
    =
    \hat{\mathbf{A}} \hat{\mathbf{c}}
\end{equation}

\noindent where $\hat{\mathbf{A}}$ is the scaled Jacobian matrix and
$\hat{\mathbf{c}}$ is a vector containing scaled coefficients.

The misfit function defined on equation~\ref{eq:misfit-unscaled} can be then
rewritten as follows:

\begin{equation}
    \phi(\hat{\mathbf{c}}) =
        \left\lVert \mathbf{d}
        - \hat{\mathbf{A}}\hat{\mathbf{c}} \right\rVert ^ 2
        + \lambda \left\lVert \hat{\mathbf{c}} \right\rVert ^2,
    \label{eq:misfit}
\end{equation}

\noindent where now $\lambda$ is a \emph{dimensionless} damping parameter.

Misfit function from equation~\ref{eq:misfit} is minimized through Cholesky
decomposition.
If the number of observation points is equal or larger to the number of
sources, then the linear problem defined on equation~\ref{eq:linear-problem} is
well- or over-determined, respectively. Therefore, the scaled coefficients that
minimize the misfit function can be obtained through \citep{menke1989}:

\begin{equation}
    \hat{\mathbf{c}} =
        \left[
            \hat{\mathbf{A}}^\text{T} \hat{\mathbf{A}}
            + \lambda^2 \mathbf{I}
        \right]\inv
        \hat{\mathbf{A}}^\text{T}
        \mathbf{d}.
\end{equation}

On the other hand, if there are more sources than observation points, the
problem is under-determined and the scaled coefficients can be obtained as
\citep{menke1989}:

\begin{equation}
    \hat{\mathbf{c}} =
    \hat{\mathbf{A}}^\text{T}
        \left[
            \hat{\mathbf{A}} \hat{\mathbf{A}}^\text{T}
            + \lambda^2 \mathbf{I}
        \right]\inv
        \mathbf{d}.
\end{equation}

Once the scaled coefficients are obtained, the sources' coefficients
$\mathbf{c}$ can be obtained by rescaling
$\hat{\mathbf{c}}$:

\begin{equation}
    \mathbf{c} = \mathbf{S}\inv \hat{\mathbf{c}}.
\end{equation}

\subsection{Gradient boosting}

Gradient boosting was first introduced by \citet{friedman2001, friedman2002} as
a method for fitting additive parametric models of the form

\begin{equation}
  d = \sum_{k=1}^K \alpha_k f(\mathbf{c}_k),
\end{equation}

\noindent where $\alpha_k$ is a scalar coefficient called the \emph{step-size}
and $f$ is a function of the parameter vector $\mathbf{c}_k$.
For linear regression problems, these additive models can be written as the
matrix equation

\begin{equation}
    \mathbf{d} = \sum_{k=1}^K \alpha_k \mathbf{A}_k \mathbf{c}_k \ .
    \label{eq:gb-linear-model}
\end{equation}

We can transform our equivalent source problem in
equation~\ref{eq:linear-problem} into an additive model by
(1) defining a set of $K$ overlapping windows of equal size that cover the
survey area and (2) defining one set of equivalent sources per
window. Then, the coefficients $\mathbf{c}_k$ relate to the equivalent sources
of the $k$-th window and $\mathbf{A}_k$ is the Jacobian matrix between
those sources and every data point.
Thus, the predicted data are a superposition of the
effects of the $K$ sets of equivalent sources.

The gradient boosting algorithm above works by fitting each component of the
additive model, one at a time, to the residuals of the previous component.
\citet{friedman2001} demonstrates that this corresponds to a steepest-descent
optimization in the so-called ``function space'', with $\alpha_k$ corresponding
to step sizes in steepest descent.
We can modify the gradient boosting algorithm of \citet{friedman2001} to
find the damped least-squares solutions for the $K$ parameters vectors
$\mathbf{c}_k$ and step sizes $\alpha_k$ in equation~\ref{eq:gb-linear-model}:

\begin{algorithm}[!h]
  \DontPrintSemicolon
  \setstretch{1.5}
  define the residual vector: $\mathbf{r}_{0} = \mathbf{d}^o$ \;
  \For{ $k = 1$ \KwTo $K$ }{

    \nllabel{alg:scale}
    $\mathbf{B}_k = \mathbf{A}_k \mathbf{S}_k\inv$ \;

    \nllabel{alg:fit}
    $
     \hat{\mathbf{m}}_k = \left[\mathbf{B}_k\trans \mathbf{W}_k \mathbf{B}_k +
     \lambda \mathbf{I} \right]\inv \mathbf{B}_k\trans \mathbf{W}_k
     \mathbf{r}_{k-1}
    $ \;

    \nllabel{alg:unscale}
    $\hat{\mathbf{c}}_k = \mathbf{S}_k\inv \hat{\mathbf{m}}_k$ \;

    \nllabel{alg:stepsize}
    $
      \alpha_k = \underset{\alpha}{\mathrm{argmin}}
      \left(
        \left\lVert
          \mathbf{r}_{k - 1} - \alpha \mathbf{A}_k \hat{\mathbf{c}}_k
        \right\rVert ^ 2
      \right)
    $ \;

    \nllabel{alg:residual}
    $\mathbf{r}_k = \mathbf{r}_{k - 1} - \alpha_k\mathbf{A}_k \hat{\mathbf{c}}_k$ \;
  }
  \BlankLine
  \setstretch{1}
  \caption{Gradient boosting solution for damped least-squares regression.}
\end{algorithm}

\citet{friedman2002} suggests that introducing randomness into the fitting
process can improve convergence of the solution. To achieve this, we randomize
the order in which the $K$ windows are used in the gradient boosting algorithm.

The $\mathbf{A}_k$ matrices have only $N \times M_k$ elements
(where $M_k$ is the number sources on the $k-$th window), which can be
considerably smaller than the $N \times M$ elements of $\mathbf{A}$.
Therefore, the gradient boosting method allows us to fit
equivalent sources models that would produce Jacobian matrices that are larger
than the available computer memory.
Furthermore, we can increase or decrease the size of the overlapping windows as
needed depending on the number of sources in the model and the available
computer memory.

We can improve the efficiency of the algorithm further by using only the $N_k$ data
points that fall within the $k$-th window for fitting the sources
(steps \ref{alg:scale} and \ref{alg:fit}),
while using all available data when calculating the step size and residuals
(steps \ref{alg:stepsize} and \ref{alg:residual}).
By doing so, we reduce the Jacobian matrix further to $N_k \times M_k$
elements.
The forward modeling operations performed in steps \ref{alg:stepsize} and
\ref{alg:residual} can be done by a summation
(TODO: ADD EQUATION)
instead of a matrix-vector product, which allows us to avoid computing and
storing the larger $N \times M_k$ Jacobian at any point.
The influence of distance sources on the solution is maintained by using the
entire set of data points when updating the residuals vector.

The step size parameter functions as scaling factor that prevents the algorithm
from worsening the fit to data points outside of the current window.
As such, it improves the converge of the solution and ensures that the
overall misfit decreases with iteration.
The step size that minimizes the misfit function in step \ref{alg:stepsize} of
the algorithm is

\begin{equation}
    \alpha_k = \frac{
      \mathbf{r}_{k-1}\trans \mathbf{A}_k \hat{\mathbf{c}}_k
    }{
      \hat{\mathbf{c}}_k\trans\mathbf{A}_k\trans \mathbf{A}_k \hat{\mathbf{c}}_k
    }.
\end{equation}

After all $\alpha_k$ step sizes and $\hat{\mathbf{c}}_k$ coefficients vectors
are estimated, we can predict the effect of the additive equivalent source
model on any point through the summation

\begin{equation}
    d(\mathbf{p}) =
    \sum\limits_{k=1}^K \sum\limits_{j=1}^{M_k}
    \alpha_k \frac{{c_k}_j}{\left\lVert \mathbf{p} - \mathbf{q}_j \right\rVert}
    \ .
\end{equation}

NOTE: This paragraph might be better in the Discussion section.

Our gradient boosting algorithm for overlapping windows is similar to the
``bootstrap inversion'' used in \citet{vonfrese1988}, which also iteratively
fits portions of an equivalent source model to the data residuals.
The key differences are:
(1) the sources in the overlapping portions of the windows are fit more than
once, allowing the algorithm to self-correct for poor solutions to any given
window;
(2) using only data points within the window when fitting the data enables the
use of larger datasets;
(3) experience shows that the step size parameter improves the convergence of
the solution, particularly when using only the data inside the windows for
fitting.

@@

\begin{table*}
    \centering
    \caption{
        Source distributions as combinations of source layouts and depth
        strategies.
    }
    \label{tab:source-distributions}
    \begin{tabular}{lccc}
        & Source below data & Block averaged sources & Grid sources \\ \hline
        Constant depth & $\checkmark$ & $\checkmark$ & $\checkmark$ \\
        Relative depth & $\checkmark$ & $\checkmark$ & $\times$     \\
        Variable depth & $\checkmark$ & $\checkmark$ & $\times$     \\
    \end{tabular}
\end{table*}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
